\frontmattersection{Acknowledgements}

\enlargethispage{2\baselineskip}

%This thesis would not have been possible without .

%My thanks to my supervisor Amos Storkey whose patience, insightful comments and 

It has been a privilege to have had Amos Storkey as my supervisor. He has unerringly been able to provide helpful insight in to any problem I came to him with and the enthusiasm and humour he has brought to all our discussions has made it a joy to work with him. I also feel very fortunate to have had the chance to collaborate with Iain Murray on the work described in Chapter \ref{ch:pseudo-marginal-methods} and I am extremely grateful to him for all the additional helpful and perceptive advice, ideas and comments he has given me during my research. My thanks also to Peggy Seri\`{e}s, Matthias Hennig and Mark van Rossum for providing helpful guidance at review meetings over the course of my research and to my examiners, Dennis Prangle and Lukasz Szpruch, for their patience in reviewing this thesis and for their useful comments and feedback.
%my secondary supervisor Peggy Seri\`{e}s, who provided helpful guidance in the early stages of my project and would have undoubtedly continued doing so if I had not diverged so much from my original proposed topic!

%I am also very grateful to Iain Murray for offering me the chance to collaborate with him on the work which is described in Chapter \ref{ch:pseudo-marginal-methods} and for all the helpful advice, ideas and comments I have received from him over the course of my project.

%My thanks also to my secondary supervisor Peggy Seri\`{e}s, who provided helpful guidance in the early stages of my project.

%My thanks also to my secondary supervisor Peggy Seri\`{e}s, who provided helpful guidance in the early stages of my project
%and would have undoubtedly continued doing so if I had not diverged so much from my original proposed topic!

I am fortunate to have been part of the Doctoral Training Centre (DTC) in Neuroinformatics and Computational Neuroscience during my studies in Edinburgh and I would like to thank the DTC executive committee members for all of the support and opportunities the DTC has provided me with. Many thanks also to Teresa Ironside, Mila Vukomanovic and Alison May-Edie for the excellent support they provided in their tenures in charge of DTC administrative duties.

I am very grateful to the funding bodies of the DTC, the Engineering and Physical Sciences Research Council, Biotechnology and Biological Sciences Research Council and Medical Research Council, for their financial support of my research and travel. My thanks also to the Scottish Informatics and Computing Science Alliance and Disney Research for providing additional travel funding.

It has been a pleasure to work with and around such an interesting, friendly and diverse group of people at the School of Informatics. My particular thanks to my fellow DTC students for many interesting and thought-provoking discussions and journal club presentations but also for the friendships that have made my time in Edinburgh so enjoyable. I am also very grateful to the other members of Amos' research group and the wider Institute for Adaptive and Neural Computation machine learning group for numerous helpful paper discussions and whiteboard brainstorm sessions.

Last but definitely not least - I could not have got to this point without the continual support and encouragement of my family and friends. My heartfelt thanks to all of you.

%for every shared moment of joy, wonder and laughter.